---
title: "R for Geospatial Workshop"
author: "Keene Morrow"
date: "11/21/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rgdal)
library(raster)
library(here)
```
# [Introduction to Raster Data](https://datacarpentry.org/r-raster-vector-geospatial/01-raster-structure/index.html)

## Raster Basics

```{r}
# Load digital surface model for Harvard Forest
# 1m GEOTIF
# UTM Zone 18
# can navigatre through subfolders by hitting tab
HARV_dsmCrop_info <- GDALinfo("data/NEON-DS-Airborne-Remote-Sensing/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif")

HARV_dsmCrop_info

# Load the GEOTIF as a raster
DSM_HARV <- raster(here::here("data", "NEON-DS-Airborne-Remote-Sensing", "NEON-DS-Airborne-Remote-Sensing", "HARV", "DSM", "HARV_dsmCrop.tif"))

DSM_HARV

# call summary of raster
summary(DSM_HARV) # using 4.31% of cells
summary(DSM_HARV, maxsamp=ncell(DSM_HARV)) # using all the cells
```

```{r}
#DSM_HARV is currently a raster... we want a data frame to plot in ggplot

DSM_HARV_df <- as.data.frame(DSM_HARV, xy = TRUE) %>%
  mutate(z = HARV_dsmCrop)
# this df has more than 2 million rows
# consider whether a raster really needs to be a data frame...

ggplot() +
  geom_raster(data = DSM_HARV_df, aes(x = x, y = y, fill = z)) +
  scale_fill_viridis_c() +
  coord_quickmap()


```

```{r}
plot(DSM_HARV) # makes a gross looking plot of the raster, but it's paster than ggplot for this purpose
```

## Coordinate Reference Systems

## Challenge
Use the output from the GDALinfo() function to find out what NoDataValue is used for our DSM_HARV dataset.

```{r}
GDALinfo("data/NEON-DS-Airborne-Remote-Sensing/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif")

```

Found in first data.frame output from GDALinfo()
NoDataValue = -9999

## Histogram of raster
```{r}
ggplot() +
    geom_histogram(data = DSM_HARV_df, aes(HARV_dsmCrop))
```

# [Plot Raster Data in R](https://datacarpentry.org/r-raster-vector-geospatial/02-raster-plot/index.html)

# [Reproject Raster Data in R](https://datacarpentry.org/r-raster-vector-geospatial/03-raster-reproject-in-r/index.html)

## Raster Projection in R

```{r}
DTM_HARV <- raster(here::here("data", "NEON-DS-Airborne-Remote-Sensing", "NEON-DS-Airborne-Remote-Sensing", "HARV", "DTM", "HARV_dtmCrop.tif")) # WGS84 using UTM Zone 18

DTM_HARV
```


```{r}
DTM_hill_HARV <- raster(here::here("data", "NEON-DS-Airborne-Remote-Sensing", "NEON-DS-Airborne-Remote-Sensing", "HARV", "DTM", "HARV_DTMhill_WGS84.tif")) # WGS84 using lat-long

crs(DTM_hill_HARV)

```

### Reprojecting the hillshade file

*Reprojecting changes the data!*
```{r}
DTM_hill_UTMZ18N_HARV <- projectRaster(DTM_hill_HARV, crs = crs(DTM_HARV))

```

```{r}
crs(DTM_hill_UTMZ18N_HARV)
extent(DTM_hill_UTMZ18N_HARV)

crs(DTM_HARV)
```

### [Raster Calculations in R](https://datacarpentry.org/r-raster-vector-geospatial/04-raster-calculations-in-r/index.html): Canopy Height Model
Creating a canopy height model by subtracting digital terrain model from digital surface model

```{r}
# double check that the projections are the same
crs(DTM_HARV)
crs(DSM_HARV)

# check that the extents are the same
extent(DTM_HARV)
extent(DSM_HARV)
# they are
# if they weren't, we'd use clip()

# and check them out visually
plot(DTM_HARV) # smoother and lower elevation
plot(DSM_HARV) # rougher and higher elevation
```
#### With direct subtraction
```{r}
# quick and dirty canopy height model
CHM <-  DSM_HARV - DTM_HARV
plot(CHM)

```
#### With overlay()
overlay() takes two rasters and a function telling it what to do with them


```{r}
CHM_ov_HARV <- overlay(DSM_HARV, DTM_HARV, fun = function(r1, r2) {return(r1 - r2)})

plot(CHM_ov_HARV)
```

#### Challenge: Explore CHM Raster Values
It’s often a good idea to explore the range of values in a raster dataset just like we might explore a dataset that we collected in the field.

1. What is the min and maximum value for the Harvard Forest Canopy Height Model (CHM_HARV) that we just created?
```{r}
min(values(CHM_ov_HARV), na.rm = TRUE)
max(values(CHM_ov_HARV), na.rm = TRUE)

```


2. What are two ways you can check this range of data for CHM_HARV?


3. What is the distribution of all the pixel values in the CHM?


4. Plot a histogram with 6 bins instead of the default and change the color of the histogram.
```{r}
CHM_df <- as.data.frame(CHM, xy = TRUE)
```

```{r}
ggplot() +
  geom_histogram(data = CHM_df,
                 aes(x = layer),
                 # bins = 6,
                 fill = "darkgreen",
                 color = "grey20") +
  theme_minimal()
```


5. Plot the CHM_HARV raster using breaks that make sense for the data. Include an appropriate color palette for the data, plot title and no axes ticks / labels.


### Export a GeoTIFF

```{r}
writeRaster(CHM_ov_HARV, "CHM_HARV.tiff",
            format="GTiff",
            overwrite=TRUE,
            NAflag=-9999)
```

### [Work With Multi-Band Rasters in R](https://datacarpentry.org/r-raster-vector-geospatial/05-raster-multi-band-in-r/index.html)

#### Getting Started with Multi-Band Data in R

```{r}
RGB_band1_HARV <- raster("data/NEON-DS-Airborne-Remote-Sensing/NEON-DS-Airborne-Remote-Sensing/HARV/RGB_Imagery/HARV_RGB_Ortho.tif")

RGB_band1_HARV_df  <- as.data.frame(RGB_band1_HARV, xy = TRUE)

ggplot() +
  geom_raster(data = RGB_band1_HARV_df,
              aes(x = x, y = y, alpha = HARV_RGB_Ortho)) + 
  coord_quickmap()
```

```{r}
RGB_stack_HARV <- stack(here::here("data", "NEON-DS-Airborne-Remote-Sensing", "NEON-DS-Airborne-Remote-Sensing", "HARV", "RGB_Imagery", "HARV_RGB_Ortho.tif"))
```

